$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-03 paper/paper.tex
Pass to the intro

We live in a concurrent world where any given state is often computed interactively by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted.
The \picalc{} acts as a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
Moreover, at the state-of-the-art, the \picalc{} features a wide plethora of types: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types \cite{K07}. This makes the \picalc{} a fully-fledged tool for the modeling and verification of concurrent and distributed systems.

@@

In the meantime, the rise of session types \cite{H93,THK94,HVK98} gave a different flavour to the linearity in the \picalc{}.
Session types are a type formalism used in communication-centric programming.
In session types, linearity ensures that a channel must be owned by exactly one communicating participant, even if the channel itself can be used multiple times, as specified by the structure of its session type.
Linearity in session types is the key ingredient that guarantees type safety.

Recent work \cite{DardhaGS12,Dardha14,DardhaGS17} pushed the linear \picalc{} into the spotlight again \cite{KPT96}, thanks to an encoding of session types into linear types.
This encoding not only has theoretical benefits in terms of expressivity of session types and reusability of theoretical results from the linear \picalc{}, but is useful at a practical level too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}.
This means that we can use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

Considering the importance of the \picalc{} and of linearity in modeling concurrent and distributed systems, in this paper we focus our research on both and go beyond.
We present for the first time a full mechanisation in Agda of a \picalc{} with linear, gradual and shared types, all under the same unified framework.
We do so by defining a specification for an algebra on multiplicities and using it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
Ultimately, we can use $\pi$ with leftovers as a unified framework for type safe distributed modeling and programming with a variety of type systems, on top of which more advanced types, theories and languages can be built.
While linearity is needed for type safety, gradual types and shared types are needed for real-world systems.

@@

Linear, gradual and shared types (identified by a value in \Idx) are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.

>>>>

We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features \emph{channel mobility}: channels themselves are first order constructs, and can therefore be transmitted.
The \picalc{} acts as a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
At the state-of-the-art, the \picalc{} features a wide plethora of types: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types \cite{K07}. This makes the \picalc{} a fully-fledged tool for the modeling and verification of concurrent and distributed systems.

@@

In the meantime, the rise of session types \cite{H93,THK94,HVK98} gave a different flavour to linearity in the \picalc{}.
Session types are a type formalism used in communication-centric programming.
In session types, linearity ensures that a channel is owned by exactly one communicating participant, even if the channel itself can be used multiple times, following the structure of its session type.
Linearity in session types is the key ingredient that guarantees type safety.

Recent work \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} into the spotlight again \cite{KPT96}, thanks to an encoding of session types into linear types.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful at a practical level too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}.
This means that we can use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types are built.

Considering the importance of the \picalc{} and of linearity in modeling concurrent and distributed systems, in this paper we focus our research on both and go beyond. \todo{}
We present the first full mechanisation in Agda of a \picalc{} with linear, gradual and shared types, all under the same unified framework.
We do so by defining a specification for an algebra on multiplicities and using it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
Ultimately, we can use this work as a unified framework for type safe distributed modeling and programming with a variety of type systems, and built more advanced types, theories and languages on top.
While linearity is needed for type safety, gradual and shared types are needed for real-world systems.

@@

Linear, gradual and shared types (identified by a value in $\Idx$) are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.
